---
title: "OZNAL - Health Outcomes Analysis - PLACES Dataset"
output: html_document
date: "2026-04-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r install_packages, eval=FALSE}
install.packages(c(
  "tidyverse", "rsample", "caret", "corrplot", "patchwork",
  "glmnet", "arm", "rpart", "rpart.plot", "e1071",
  "gbm", "earth", "car", "ggcorrplot", "randomForest"
))
```

```{r libraries}
library(tidyverse)
library(rsample)
library(caret)
library(corrplot)
library(patchwork)
library(glmnet)
library(arm)
library(randomForest)
library(e1071)
library(gbm)
library(earth)
library(car)
library(ggcorrplot)
library(conflicted)

conflicts_prefer(dplyr::filter) 
conflicts_prefer(tidyr::extract) 
conflicts_prefer(dplyr::select)

set.seed(42)
```

```{r load_data}
df <- read_csv("PLACES__ZCTA_Data_(GIS_Friendly_Format),_2025_release_20260423.csv")
```
## Hypotheses

We target **stroke prevalence** ('Stroke_m') as it is a severe, often fatal outcome with well-documented, modifiable risk factors. We hypothesize that:

- Metabolic conditions (diabetes, coronary heart disease) are strongest predictors of area-level stroke prevalence.
- Behavioral factors (smoking, physical inactivity, obesity, sleep deprivation) contribute independently.
- After controlling for metabolic conditions, behavioral predictors explain residual variance.

**Scenario 1 — Baseline models:** Linear Regression, kNN Regression, Random Forest Regression.

**Scenario 2 — Parametric regression:** Bayesian Linear Regression, Ridge Regression, Lasso Regression.

**Scenario 2 — Nonparametric / flexible regression:** Support Vector Regression, Gradient Boosting Regression, MARS.

## Exploratory data analysis
### Inspecting data 

We need to find out the size of the dataset and the names of columns:

```{r size, include=TRUE}
dim(df)
```
```{r colnames, include=TRUE}
colnames(df)
```
We can see that there are columns ZCTA5 - this is unique id value for a row, which identifies area from which values of parameters are measured. There are columns for population and population over 18, and also geolocation. Other columns which have the main values are paired up, the one ending with \_crudePrev show average of that value in an area, and the ones ending with \_CI show two values, which mean that there is 95% confidence that value is somewhere in this interval.

### Data formats

```{r structure, include=TRUE}
str(df)
```

We need to change population values to integers instead of strings. We can also see that some columns have char format instead of number. We need to find them and change appropriately.

```{r population_format, include=TRUE}
df <- df %>%
  mutate(across(c(TotalPopulation, TotalPop18plus), ~ {
    clean_val <- gsub("[^0-9]", "", .x)
    as.numeric(clean_val)
  }))
```

We also need to solve the problem with CI columns. The 95% CI columns arrive as character signs `"(low, high)"`. 
```{r ci_columns, include=TRUE}
ci_cols <- names(df)[grep("_Crude95CI", names(df))]

for (col in ci_cols) {
  width_name <- paste0(col, "_width")
  high_name <- paste0(col, "_high")
  
  clean_vec <- str_remove_all(df[[col]], "[()\\s]")
  split_matrix <- str_split_fixed(clean_vec, ",", 2)
  df[[width_name]] <- as.numeric(as.numeric(split_matrix[, 2]) - as.numeric(split_matrix[, 1]))
  df[[high_name]] <- as.numeric(split_matrix[, 2])
  
  df <- df %>%
    relocate(all_of(width_name), .after = !!sym(col)) %>%
    select(-!!sym(col))
}
```

We calculate the range of values for 95 CI percentage. To keep min and max value would not be very useful, as that infomation would highly correlate with crudePrev average values, but as range, it could give us information about certainty if the range is small enough and would increase the strength of the predictions.

```{r fix_col_char, include=TRUE}
cols_crude_char <- df %>%
  select(ends_with("CrudePrev") & where(is.character)) %>%
  colnames()

print(cols_crude_char)
```

```{r fix_char}
df <- df %>%
  mutate(across(all_of(cols_crude_char), ~ {
    as.numeric(gsub(",", ".", .x))
  }))

str(df %>% select(all_of(cols_crude_char)))
```

We can also see that columns which were not included in the previous part, that they lost commas.

```{r fix_col_num, include=TRUE}
cols_crude_num <- df %>%
  select(ends_with("CrudePrev") & where(is.numeric)) %>%
  colnames()

print(cols_crude_num)
```

```{r fix_num, include=TRUE}
walk(cols_crude_num, function(col_name) {
  high_name <- paste0(str_remove(col_name, "CrudePrev"), "Crude95CI_high")
  
  
  if (high_name %in% colnames(df)) {
    df <<- df %>%
      mutate(!!sym(col_name) := if_else(
        !is.na(!!sym(col_name)) & !is.na(!!sym(high_name)) & !!sym(col_name) > !!sym(high_name),
        !!sym(col_name) / 10,
        !!sym(col_name)
      ))
  }
})
df <- df %>%
  select(-ends_with("high"))

```

Extracting geolocation.

```{r geolocation_col, include=TRUE}

df <- df %>%
  separate_wider_regex(
    cols = Geolocation,
    patterns = c(
      ".*\\(",                 
      Longitude = "-?\\d+\\.\\d+", 
      " ",                     
      Latitude = "-?\\d+\\.\\d+",
      "\\)"                   
    ),
    too_few = "align_start"
  ) %>%
  mutate(
    Longitude = as.numeric(Longitude),
    Latitude = as.numeric(Latitude)
  )
```

```{r str_repeat, include=TRUE}
str(df)
```

### Duplicates

Because we know that ZCTA5 acts as id, it is necessary to check only this column for duplicates.

```{r duplicates, include=TRUE}
duplicates <- df %>% 
  filter(duplicated(ZCTA5))
duplicates
```

We can see that in this dataset there are no duplicate rows.

### Missing values

We need to check how many missing values is in the dataset

```{r missing, include=TRUE}

df %>%
  summarise(across(everything(), ~ sum(is.na(.x)))) %>%
  pivot_longer(everything(), names_to = "column_name", values_to = "na_count") %>%
  mutate(
    percentage = (na_count / nrow(df)) * 100
  ) %>%
  print(n = Inf)

df %>%
  mutate(na_per_row = rowSums(is.na(.))) %>%
  count(na_per_row) %>%
  mutate(percentage = (n / sum(n)) * 100)

```
```{r complete_cases}
df %>%
  filter(complete.cases(.)) %>%
  nrow()
```

We can see that majority of columns have 7.8% values missing and there are 7 pairs of columns which have 25.5% of missing values, and also that there are almost 75% of rows which have no missing values.

```{r missing_visualization}
df_viz <-df %>%
  mutate(
    na_count = rowSums(is.na(.)),
    na_percent = rowMeans(is.na(.)) * 100
  )
ggplot(df_viz, aes(x = Longitude, y = Latitude, color = na_percent)) +
  geom_point(alpha = 0.6, size = 1) +
  scale_color_viridis_c(option = "plasma") + 
  theme_minimal() +
  labs(
    title = "Spatial Distribution of Missing Values",
    subtitle = "Points colored by number of missing values percentage per row",
    x = "Longitude",
    y = "Latitude",
    color = "NA percent"
  )
```

We can see that the missing values are influenced by states from which they are collected. For this reason, we will not remove all the missing rows, because for many of states we would completely lose all the information. If later there are features we need missing, we will deal with it.

### Column names

We need to rename columns as some names are not informative enough.

```{r colnames2}
colnames(df)
```

```{r rename_columns}
df <- df %>%
  rename(
    Insurance_lack_m = ACCESS2_CrudePrev,
    Insurance_lack_w = ACCESS2_Crude95CI_width,
    Arthritis_m = ARTHRITIS_CrudePrev,
    Arthritis_w = ARTHRITIS_Crude95CI_width,
    Drinking_m = BINGE_CrudePrev,
    Drinking_w = BINGE_Crude95CI_width,
    HBP_m = BPHIGH_CrudePrev,
    HBP_w = BPHIGH_Crude95CI_width,
    HBP_med_m = BPMED_CrudePrev,
    HBP_med_w = BPMED_Crude95CI_width,
    Cancer_m = CANCER_CrudePrev,
    Cancer_w = CANCER_Crude95CI_width,
    Asthma_m = CASTHMA_CrudePrev,
    Asthma_w = CASTHMA_Crude95CI_width,
    CHD_m = CHD_CrudePrev,
    CHD_w = CHD_Crude95CI_width,
    Checkup_m = CHECKUP_CrudePrev,
    Checkup_w = CHECKUP_Crude95CI_width,
    Chol_screen_m = CHOLSCREEN_CrudePrev,
    Chol_screen_w = CHOLSCREEN_Crude95CI_width,
    Colon_screen_m = COLON_SCREEN_CrudePrev,
    Colon_screen_w = COLON_SCREEN_Crude95CI_width,
    COPD_m = COPD_CrudePrev,
    COPD_w = COPD_Crude95CI_width,
    Smoking_m = CSMOKING_CrudePrev,
    Smoking_w = CSMOKING_Crude95CI_width,
    Dental_m = DENTAL_CrudePrev,
    Dental_w = DENTAL_Crude95CI_width,
    Depression_m = DEPRESSION_CrudePrev,
    Depression_w = DEPRESSION_Crude95CI_width,
    Diabetes_m = DIABETES_CrudePrev,
    Diabetes_w = DIABETES_Crude95CI_width,
    General_hlth_m = GHLTH_CrudePrev,
    General_hlth_w = GHLTH_Crude95CI_width,
    High_chol_m = HIGHCHOL_CrudePrev,
    High_chol_w = HIGHCHOL_Crude95CI_width,
    No_PA_m = LPA_CrudePrev,
    No_PA_w = LPA_Crude95CI_width,
    Mammography_m = MAMMOUSE_CrudePrev,
    Mammography_w = MAMMOUSE_Crude95CI_width,
    Mental_distress_m = MHLTH_CrudePrev,
    Mental_distress_w = MHLTH_Crude95CI_width,
    Obesity_m = OBESITY_CrudePrev,
    Obesity_w = OBESITY_Crude95CI_width,
    Physical_distress_m = PHLTH_CrudePrev,
    Physical_distress_w = PHLTH_Crude95CI_width,
    Short_sleep_m = SLEEP_CrudePrev,
    Short_sleep_w = SLEEP_Crude95CI_width,
    Stroke_m = STROKE_CrudePrev,
    Stroke_w = STROKE_Crude95CI_width,
    Teeth_lost_m = TEETHLOST_CrudePrev,
    Teeth_lost_w = TEETHLOST_Crude95CI_width,
    Hearing_m = HEARING_CrudePrev,
    Hearing_w = HEARING_Crude95CI_width,
    Vision_m = VISION_CrudePrev,
    Vision_w = VISION_Crude95CI_width,
    Cognition_m = COGNITION_CrudePrev,
    Cognition_w = COGNITION_Crude95CI_width,
    Mobility_m = MOBILITY_CrudePrev,
    Mobility_w = MOBILITY_Crude95CI_width,
    Selfcare_m = SELFCARE_CrudePrev,
    Selfcare_w = SELFCARE_Crude95CI_width,
    Indep_living_m = INDEPLIVE_CrudePrev,
    Indep_living_w = INDEPLIVE_Crude95CI_width,
    Disability_m = DISABILITY_CrudePrev,
    Disability_w = DISABILITY_Crude95CI_width,
    Loneliness_m = LONELINESS_CrudePrev,
    Loneliness_w = LONELINESS_Crude95CI_width,
    Food_stamp_m = FOODSTAMP_CrudePrev,
    Food_stamp_w = FOODSTAMP_Crude95CI_width,
    Food_insecure_m = FOODINSECU_CrudePrev,
    Food_insecure_w = FOODINSECU_Crude95CI_width,
    House_insecure_m = HOUSINSECU_CrudePrev,
    House_insecure_w = HOUSINSECU_Crude95CI_width,
    Utility_threat_m = SHUTUTILITY_CrudePrev,
    Utility_threat_w = SHUTUTILITY_Crude95CI_width,
    Transport_lack_m = LACKTRPT_CrudePrev,
    Transport_lack_w = LACKTRPT_Crude95CI_width,
    Support_lack_m = EMOTIONSPT_CrudePrev,
    Support_lack_w = EMOTIONSPT_Crude95CI_width
  )
```

### Visualizations
```{r histogram}
df %>%
  select(where(is.numeric) & ends_with("_m")) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = value)) +
  geom_histogram(fill = "steelblue", color = "white", bins = 30) +
  facet_wrap(~variable, scales = "free") + 
  theme_minimal() +
  labs(title = "Histograms of Numeric Variables", x = "Value", y = "Frequency")
```
```{r scatter_pairs}
df %>%
  pivot_longer(
    cols = ends_with(c("_m", "_w")),
    names_to = c("variable", ".value"),
    names_pattern = "(.*)_(.*)"
  ) %>%
  filter(is.finite(m), is.finite(w)) %>%
  ggplot(aes(x = m, y = w)) +
  geom_point(alpha = 0.08, size = 0.3) +
  facet_wrap(~variable, scales = "free", ncol = 6) +
  labs(
    x = "Mean (Average)",
    y = "Width (Spread)",
    title = "Scatterplots of Mean vs. Width by Variable"
  ) +
  theme_bw() +
  theme(
    strip.text = element_text(size = 8),
    axis.text = element_text(size = 6)
  )
```
```{r scatter_with_predictors}
target <- "Stroke_m"

df %>%
  # 1. Select the target and all other '_m' columns
  select(all_of(target), ends_with("_m")) %>%
  # 2. Pivot everything EXCEPT the target into a long format
  pivot_longer(
    cols = -all_of(target),
    names_to = "comparison_var",
    values_to = "comparison_val"
  ) %>%
  # 3. Plot target on X and the other values on Y
  ggplot(aes(y = .data[[target]], x = comparison_val)) +
  geom_point(alpha = 0.01) +
  facet_wrap(~comparison_var, scales = "free_y") +
  labs(
    y = paste("Reference:", target),
    x = "Comparison Mean Values",
    title = paste("Comparing", target, "to all other Means")
  ) +
  theme_minimal()
```

```{r boxplots}
df %>%
  select(where(is.numeric) & ends_with("_m")) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
  ggplot(aes(y = value, x = "")) +
  geom_boxplot(fill = "orange", outlier.color = "red", outlier.shape = 1) +
  facet_wrap(~variable, scales = "free") +
  theme_minimal() +
  labs(title = "Boxplots of Numeric Variables", y = "Value", x = NULL)

```

## Feature Selection
### Correlation with Stroke_m
```{r stroke_correlations}
corr_tbl <- df %>%
  select(ends_with("_m") & where(is.numeric)) %>%
  summarise(across(-Stroke_m,
                   ~ cor(.x, Stroke_m, use = "complete.obs"))) %>%
  pivot_longer(everything(), names_to = "Feature", values_to = "r") %>%
  mutate(abs_r = abs(r)) %>%
  arrange(desc(abs_r))

print(corr_tbl, n = Inf)

corr_tbl %>%
  slice_head(n = 20) %>%
  ggplot(aes(x = reorder(Feature, abs_r), y = r,
             fill = r > 0)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "steelblue", "FALSE" = "tomato"),
                    labels = c("Negative", "Positive"), name = NULL) +
  theme_minimal() +
  labs(title = "Top 20 features by |r| with Stroke_m",
       x = NULL, y = "Pearson r")
```
### Feature exclusion rationale
We exclude features that are **consequences** of stroke rather than causes - including them would create **data leakage** (the model would learn from outcomes, not risk factors):
| Excluded feature | Reason |
|---|---|
| `Mobility_m` | r = 0.957 — mobility impairment is a direct consequence of stroke |
| `Selfcare_m`, `Indep_living_m`, `Disability_m` | Downstream disability outcomes |
| `Hearing_m`, `Vision_m`, `Cognition_m` | Sensory/cognitive impairments caused by stroke |
| `Physical_distress_m`, `General_hlth_m` | General health status downstream of stroke |
| `Teeth_lost_m`, `Arthritis_m` | Comorbidities with high confounding, not causal |
| `COPD_m` | High VIF (28) — collinear with Diabetes + CHD |
| `HBP_m` | VIF = 18 when combined with Diabetes + CHD — dropped to reduce multicollinearity; CHD captures shared vascular pathway |

**Retained features** are all causally upstream, modifiable risk factors with domain support:
| Feature | r with Stroke | Category | Rationale |
|---|---|---|---|
| `Diabetes_m` | 0.944 | Metabolic | Primary modifiable metabolic risk |
| `CHD_m` | 0.914 | Cardiovascular | Shared vascular pathology with stroke |
| `No_PA_m` | 0.787 | Behavioral | Lack of physical activity — independently modifiable |
| `Smoking_m` | 0.728 | Behavioral | Direct vascular damage |
| `Obesity_m` | 0.636 | Metabolic/Behavioral | Modifiable, drives diabetes and HBP |
| `Short_sleep_m` | 0.399 | Behavioral | Sleep deprivation increases stroke risk |

### Collinearity check - full correlation matrix of selected features
```{r corr_matrix}
selected_for_corr <- c(
  "Stroke_m",
  "Diabetes_m",
  "CHD_m",
  "No_PA_m",
  "Smoking_m",
  "Obesity_m",
  "Short_sleep_m"
)

df %>%
  select(all_of(selected_for_corr)) %>%
  na.omit() %>%
  cor() %>%
  ggcorrplot(lab = TRUE, lab_size = 3,
             colors = c("tomato", "white", "steelblue"),
             title = "Correlation Matrix - selected features + target")
```
### VIF check
```{r selected_corr_matrix}
vif_data <- df %>%
  select(Stroke_m, Diabetes_m, CHD_m, No_PA_m,
         Smoking_m, Obesity_m, Short_sleep_m) %>%
  na.omit()

vif_vals <- lm(Stroke_m ~ Diabetes_m + CHD_m + No_PA_m +
                 Smoking_m + Obesity_m + Short_sleep_m,
               data = vif_data) %>%
  vif()

vif_df <- data.frame(Feature = names(vif_vals), VIF = round(vif_vals, 2)) %>%
  mutate(Status = case_when(
    VIF > 10 ~ "Severe (>10)",
    VIF > 5  ~ "Moderate (5-10)",
    TRUE     ~ "Acceptable (<5)"
  ))

print(vif_df)

ggplot(vif_df, aes(x = reorder(Feature, VIF), y = VIF, fill = Status)) +
  geom_col() +
  geom_hline(yintercept = 5,  linetype = "dashed", color = "orange") +
  geom_hline(yintercept = 10, linetype = "dashed", color = "red") +
  coord_flip() +
  scale_fill_manual(values = c("Acceptable (<5)" = "steelblue",
                                "Moderate (5-10)" = "orange",
                                "Severe (>10)"    = "red")) +
  theme_minimal() +
  labs(title = "VIF for selected predictors",
       subtitle = "Orange dashed = 5 threshold, Red dashed = 10 threshold",
       x = NULL, y = "VIF")
```
`Diabetes_m` (VIF ≈ 8.0) and `No_PA_m` (VIF ≈ 6.1) are in the moderate range — no predictor exceeds 10, which is the standard threshold for severe multicollinearity (ISLR). Ridge and Lasso regression in Scenario 2 will further mitigate this through regularization.

## Data preprocessing for modeling
### Select and clean columns

```{r model_cols}
model_cols <- c("Stroke_m", "Diabetes_m", "CHD_m",
                "No_PA_m", "Smoking_m", "Obesity_m", "Short_sleep_m")

df_raw_model <- df %>%
  select(all_of(model_cols)) %>%
  na.omit()

cat("Rows before outlier removal:", nrow(df_raw_model), "\n")
```
### Outliers removal - 3xIQR rule
```{r outlier_removal}
remove_outliers_iqr <- function(data, cols, k = 3) {
  mask <- rep(TRUE, nrow(data))
  for (col in cols) {
    q <- quantile(data[[col]], c(0.25, 0.75), na.rm = TRUE)
    iqr <- q[2] - q[1]
    lower <- q[1] - k * iqr
    upper <- q[2] + k * iqr
    mask <- mask & (data[[col]] >= lower) & (data[[col]] <= upper)
  }
  data[mask, ]
}

df_model <- remove_outliers_iqr(df_raw_model, model_cols)

cat("Rows after outlier removal: ", nrow(df_model), "\n")
cat("Removed:                    ", nrow(df_raw_model) - nrow(df_model),
    sprintf("(%.2f%%)\n", 100 * (nrow(df_raw_model) - nrow(df_model)) / nrow(df_raw_model)))
```
```{r outlier_comparison}
bind_rows(
  df_raw_model %>% mutate(stage = "Before"),
  df_model %>% mutate(stage = "After")
) %>%
  pivot_longer(-stage, names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = stage, y = value, fill = stage)) +
  geom_boxplot(outlier.size = 0.3) +
  facet_wrap(~variable, scales = "free") +
  theme_minimal() +
  labs(title = "Distributions before and after 3xIQR outlier removal",
       x = NULL, fill = NULL)
```
### Skewness check
```{r skewness}
skew_tbl <- df_model %>%
  summarise(across(everything(),
                   ~ round(e1071::skewness(.x, na.rm = TRUE), 3))) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Skewness") %>%
  mutate(Interpretation = case_when(
    abs(Skewness) < 0.5 ~ "Approximately symetric",
    abs(Skewness) < 1.0 ~ "Moderate skew",
    TRUE ~ "High skew"
  ))

print(skew_tbl)
```
All variables show skewness in the range [−0.5, 0.7] after outlier removal — approximately symmetric, acceptable for linear regression residuals. Log-transforming the target would actually worsen skewness (log-skew ≈ −2.2 vs raw skew ≈ 0.27), so we keep the raw scale.

## Train Test split
```{r test_train_split}
set.seed(42)

data_split <- initial_split(df_model, prop = 0.8)
train_data <- training(data_split)
test_data <- testing(data_split)

cat("Train:", nrow(train_data), "\n")
cat("Test:", nrow(test_data), "\n")

formula_stroke <- Stroke_m ~ Diabetes_m + CHD_m + No_PA_m + Smoking_m + Obesity_m + Short_sleep_m


```
```{r pairplot}
pairs(df_model,
      pch = 19, cex = 0.03, col = "steelblue",
      main = "Pairplot - modeling features after preprocessing")
```
The pairplot indicates that the selected predictors are suitable for modeling `Stroke_m`. The clearest positive relationships are visible for `Diabetes_m` and `CHD_m`, which supports their role as the strongest predictors of stroke prevalence. Behavioral variables such as `No_PA_m`, `Smoking_m`, `Obesity_m`, and `Short_sleep_m` also show positive associations, although with more variability.

Several predictors are correlated with each other, especially diabetes, coronary heart disease, physical inactivity, and obesity. This suggests moderate multicolinearity, which is expected in health data and is evaluated using VIF. Since the relationships are mostly monotonic and approximately linear, linear models are appropriate, while the visible scatter and mild nonlinearity also justify testing flexible nonparametric models.

# Scenario 1
## 1. Linear regression

#### Conditions
To use linear regression, we need to make sure that there are linear relationships between target and predictors, there should homoscedasticity, no extreme outliers, there should be no severe multicollinearity, feature scaling might help.

```{r skewness_p}
skewness_df <- df_model %>%
  summarise(across(everything(), ~ skewness(.x, na.rm = TRUE))) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Skewness") %>%
  arrange(desc(abs(Skewness)))
print(skewness_df)

```


```{r homeost}
lm_model <- lm(formula_stroke, data = train_data)

summary(lm_model)
confint(lm_model)

lm_preds <- predict(lm_model, newdata = test_data)

lm_perf <- postResample(
  pred = lm_preds,
  obs = test_data$Stroke_m
)

print(lm_perf)

par(mfrow = c(2, 2))
plot(lm_model)
par(mfrow = c(1, 1))
```

## 2. KNN
```{r knn}
knn_ctrl <- trainControl(method = "cv", number = 5)

knn_model <- train(
  formula_stroke,
  data = train_data,
  method = "knn",
  preProcess = c("center", "scale"),
  tuneGrid = expand.grid(k = c(3, 5, 7, 10, 15, 20, 25, 30, 40, 50)),
  trControl = knn_ctrl,
  metric = "RMSE"
)

print(knn_model)
plot(knn_model)

knn_preds <- predict(knn_model, newdata = test_data)

knn_perf <- postResample(
  pred = knn_preds,
  obs = test_data$Stroke_m
)

print(knn_perf)

ggplot(data.frame(obs = test_data$Stroke_m, pred = knn_preds),
       aes(x = obs, y = pred)) +
  geom_point(alpha = 0.15, size = 0.6, color = "darkgreen") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(
    title = "kNN Regression: Actual vs Predicted",
    x = "Actual Stroke %",
    y = "Predicted Stroke %"
  )
```

### 3. Random forest
```{r randomforest}
set.seed(42)

rf_grid <- expand.grid(
  mtry = c(2, 3, 4)
)

rf_model <- train(
  formula_stroke,
  data = train_data,
  method = "rf",
  tuneGrid = rf_grid,
  trControl = trainControl(method = "cv", number = 5),
  metric = "RMSE",
  ntree = 500,
  importance = TRUE
)

print(rf_model)
plot(rf_model)

rf_preds <- predict(rf_model, newdata = test_data)

rf_perf <- postResample(
  pred = rf_preds,
  obs = test_data$Stroke_m
)

print(rf_perf)

rf_importance <- varImp(rf_model)
print(rf_importance)

plot(rf_importance, main = "Random Forest: Variable Importance")

ggplot(data.frame(obs = test_data$Stroke_m, pred = rf_preds),
       aes(x = obs, y = pred)) +
  geom_point(alpha = 0.15, size = 0.6, color = "darkblue") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(
    title = "Random Forest: Actual vs Predicted",
    x = "Actual Stroke %",
    y = "Predicted Stroke %"
  )
```


## Model comparison
```{r model_comparison}
scenario1_results <- data.frame(
  Model = c(
    "Linear Regression",
    "kNN Regression",
    "Random Forest Regression"
  ),
  RMSE = c(
    lm_perf["RMSE"],
    knn_perf["RMSE"],
    rf_perf["RMSE"]
  ),
  Rsquared = c(
    lm_perf["Rsquared"],
    knn_perf["Rsquared"],
    rf_perf["Rsquared"]
  ),
  MAE = c(
    lm_perf["MAE"],
    knn_perf["MAE"],
    rf_perf["MAE"]
  )
) %>%
  mutate(
    RMSE = as.numeric(RMSE),
    Rsquared = as.numeric(Rsquared),
    MAE = as.numeric(MAE)
  ) %>%
  arrange(RMSE)

print(scenario1_results)
```
Scenario 1 provides the first comparison of baseline model families on the same prepared dataset. Linear Regression is used as an interpretable baseline, kNN as a distance-based model requiring scaling, and Random Forest as a tree-ensemble model that can capture nonlinear relationships and interactions.

# Scenario 2
Parametric models assume a **fixed linear functional form**: Stroke_m ≈ β₀ + β₁·Diabetes + β₂·CHD + … + ε. They estimate a finite, fixed number of coefficients. This makes them interpretable and statistically efficient when assumptions hold.
## Parametric models
### 1. Bayesian Linear Regression
#### Assumptions
Bayesian Linear Regression has the same basic structure as ordinary linear regression, so we checked the same assumptions. The pairplot showed approximately linear positive relationships between `Stroke_m` and the selected predictors, especially `Diabetes_m` and `CHD_m`. Observations represent separate ZCTA areas, so independence is assumed by the data structure. VIF analysis showed moderate multicollinearity between some health-related predictors, but no VIF exceeded 10, so severe multicollinearity was not present. Extreme outliers were reduced using the conservative 3×IQR rule. Residual diagnostics were inspected using the standard linear model diagnostic plots; minor deviations from perfect normality or constant variance are acceptable here because the main objective is prediction, evaluated by test-set RMSE, MAE and R².

```{r bayesian_fit}
bayes_model <- bayesglm(
  formula_stroke,
  data = train_data,
  family = gaussian()
)
summary(bayes_model)
```
```{r bayesian_diagnostics}
lm_diag <- lm(formula_stroke, data = train_data)

par(mfrow = c(2, 2))
plot(lm_diag)
par(mfrow = c(1, 1))
```
```{r bayesian_predict}
bayes_preds <- predict(bayes_model, newdata = test_data)

ggplot(data.frame(obs = test_data$Stroke_m, pred = bayes_preds),
       aes(x = obs, y = pred)) +
  geom_point(alpha = 0.15, size = 0.6, color = "purple") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(title = "Bayesian Linear Regression: Actual vs Predicted",
       x = "Actual Stroke %", y = "Predicted Stroke %")

bayes_perf <- postResample(pred = bayes_preds, obs = test_data$Stroke_m)
bayes_perf
```
### 2. Ridge Regression
#### Assumptions

Ridge adds an **L2 penalty** (λ·Σβⱼ²) to the OLS loss function. This shrinks all coefficients toward zero simultaneously, which directly addresses the moderate multicollinearity we observed between `Diabetes_m`, `No_PA_m`, and `CHD_m`. It trades a small increase in bias for potentially large reduction in variance.

Assumptions:
- **Linearity**, **independence**, **homoscedasticity**, **normality of residuals** — same as OLS.
- **Feature scaling required** — the L2 penalty treats all coefficients equally; without standardization, variables on different scales would be penalized unequally. We set `standardize = TRUE`.
- **λ** controls regularization strength. λ = 0 reduces to OLS; larger λ → stronger shrinkage. We select λ by **10-fold cross-validation** minimizing MSE.
- Coefficients approach but never reach exactly zero — Ridge retains all predictors.

```{r ridge_fit}
X_train <- model.matrix(formula_stroke, data = train_data)[, -1]
y_train <- train_data$Stroke_m
X_test  <- model.matrix(formula_stroke, data = test_data)[, -1]

ridge_cv <- cv.glmnet(
  X_train, y_train,
  alpha = 0,
  standardize = TRUE,
  nfolds = 10
)

plot(ridge_cv)
title("Ridge: CV MSE vs log(λ)", line = 3)

cat("Best lambda (min CV error):", round(ridge_cv$lambda.min, 5), "\n")
cat("Best lambda (1-SE rule):   ", round(ridge_cv$lambda.1se, 5), "\n")
```

```{r ridge_coefs}
ridge_coef_mat <- as.matrix(coef(ridge_cv, s = "lambda.min"))

ridge_coef_df <- data.frame(
  Term = rownames(ridge_coef_mat),
  Coefficient = as.numeric(ridge_coef_mat[, 1])
) %>%
  filter(Term != "(Intercept)") %>%
  arrange(desc(abs(Coefficient)))

print(ridge_coef_df)

ggplot(ridge_coef_df, aes(x = reorder(Term, abs(Coefficient)),
                          y = Coefficient,
                          fill = Coefficient > 0)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "steelblue", "FALSE" = "tomato")) +
  theme_minimal() +
  labs(
    title = "Ridge: Coefficient magnitudes at lambda.min",
    x = NULL,
    y = "Coefficient")
```
```{r ridge_predict}
ridge_preds <- as.numeric(predict(ridge_cv, s = "lambda.min", newx = X_test))

ggplot(data.frame(obs = test_data$Stroke_m, pred = ridge_preds),
       aes(x = obs, y = pred)) +
  geom_point(alpha = 0.15, size = 0.6, color = "darkred") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(title = "Ridge Regression: Actual vs Predicted",
       x = "Actual Stroke %", y = "Predicted Stroke %")
ridge_perf <- postResample(pred = ridge_preds, obs = test_data$Stroke_m)
ridge_perf
```
### 3. Lasso Regression
#### Assumptions

Lasso adds an **L1 penalty** (λ·Σ|βⱼ|) instead of L2. The key difference: the L1 penalty produces **sparse solutions** — it can set some coefficients to exactly zero, performing automatic feature selection. This is particularly valuable here because `Short_sleep_m` has the weakest correlation with Stroke_m (r = 0.40) and the Lasso can shrink it to zero if it does not contribute independently.

Assumptions:
- **Linearity**, **independence**, **homoscedasticity**, **normality of residuals** — same as Ridge.
- **Feature scaling required** — same reason as Ridge.
- **λ** selected by **10-fold cross-validation**.
- Unlike Ridge, Lasso handles collinearity by selecting one feature from a correlated group and zeroing out the others — useful given the `Diabetes_m`/`No_PA_m` correlation.

```{r lasso_fit}
lasso_cv <- cv.glmnet(
  X_train, y_train,
  alpha = 1,
  standardize = TRUE,
  nfolds = 10
)

plot(lasso_cv)
title("Lasso: CV MSE vs log(λ)", line = 3)

cat("Best lambda (min CV error):", round(lasso_cv$lambda.min, 5), "\n")
cat("Best lambda (1-SE rule):", round(lasso_cv$lambda.1se, 5), "\n")
```
```{r lasso_coefs}
lasso_coef_mat <- as.matrix(coef(lasso_cv, s = "lambda.min"))

lasso_coef_df <- data.frame(
  Term = rownames(lasso_coef_mat),
  Coefficient = as.numeric(lasso_coef_mat[, 1])
) %>%
  filter(Term != "(Intercept)") %>%
  arrange(desc(abs(Coefficient)))

print(lasso_coef_df)

# Which predictors did Lasso zero out?
zeroed <- lasso_coef_df %>%
  filter(abs(Coefficient) < 1e-8) %>%
  pull(Term)

if (length(zeroed) > 0) {
  cat("Lasso zeroed out:", paste(zeroed, collapse = ", "), "\n")
} else {
  cat("Lasso retained all predictors at lambda.min\n")
}

ggplot(lasso_coef_df, aes(x = reorder(Term, abs(Coefficient)),
                          y = Coefficient,
                          fill = Coefficient > 0)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "steelblue", "FALSE" = "tomato")) +
  theme_minimal() +
  labs(
    title = "Lasso: Coefficient magnitudes at lambda.min",
    x = NULL,
    y = "Coefficient",
    subtitle = "Zero coefficients = excluded by Lasso"
  )
```
```{r lasso_predict}
lasso_preds <- as.numeric(predict(lasso_cv, s = "lambda.min", newx = X_test))

ggplot(data.frame(obs = test_data$Stroke_m, pred = lasso_preds),
       aes(x = obs, y = pred)) +
  geom_point(alpha = 0.15, size = 0.6, color = "darkgreen") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(title = "Lasso Regression: Actual vs Predicted",
       x = "Actual Stroke %", y = "Predicted Stroke %")

lasso_perf <- postResample(pred = lasso_preds, obs = test_data$Stroke_m)
lasso_perf
```
## Nonparametric models
Nonparametric models make **no assumption about the functional form** of the relationship. They allow the data to determine the structure of the prediction surface, enabling them to capture non-linearity and interactions. This matters here because the scatter plots show mild curvature in the Diabetes–Stroke relationship.

### 1. Support Vector Regression (SVR)
SVR finds the flattest function that fits training data within a tolerance tube of width ε around predictions (the "epsilon-insensitive tube"). Only observations outside this tube (support vectors) influence the fit.

- **Feature scaling required** — SVR is distance-based; unscaled variables would dominate.
- **No distributional assumptions** on response or predictors.
- **RBF kernel** (Radial Basis Function) maps data to an infinite-dimensional space, enabling non-linear relationships without specifying a form. The kernel choice implicitly assumes local smoothness.
- **Independence** of observations — assumed.
- **No multicollinearity restriction** — but collinear features waste capacity.
- **Hyperparameters tuned via cross-validation:**
  - **C (cost)**: trade-off between margin width and training error. Higher C → less regularization.
  - **σ (sigma/gamma)**: RBF bandwidth. High σ → local influence (risk of overfitting); low σ → global.
  - **ε (epsilon)**: insensitive tube half-width — we use `method = "svmRadial"` which tunes C and sigma.

```{r svr_fit}
svr_ctrl <- trainControl(method = "cv", number = 5)

svr_grid_tuned <- expand.grid(
  C = c(25, 50, 75, 100),
  sigma = c(0.03, 0.05, 0.07)
)

svr_model_tuned <- train(
  formula_stroke,
  data = train_data,
  method = "svmRadial",
  preProcess = c("center", "scale"),
  tuneGrid = svr_grid_tuned,
  trControl = svr_ctrl,
  metric = "RMSE"
)

print(svr_model_tuned)
plot(svr_model_tuned)
```

```{r svr_predict}
svr_preds_tuned <- predict(svr_model_tuned, newdata = test_data)

ggplot(data.frame(obs = test_data$Stroke_m, pred = svr_preds_tuned),
       aes(x = obs, y = pred)) +
  geom_point(alpha = 0.15, size = 0.6, color = "firebrick") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(
    title = "Tuned SVR: Actual vs Predicted",
    x = "Actual Stroke %",
    y = "Predicted Stroke %"
  )

svr_perf_tuned <- postResample(
  pred = svr_preds_tuned,
  obs = test_data$Stroke_m
)

svr_perf_tuned
```

### 2. Gradient Boosting Regression
#### Assumptions
- **No linearity, normality, or homoscedasticity assumptions.**
- **No feature scaling required** — tree-based splits are scale-invariant.
- **Independence** of observations — assumed.
- **Robust to outliers** — tree splits are order-based.
- **Can capture non-linear patterns and interactions** automatically and more flexibly than a single tree.
- **Prone to overfitting** without regularization — controlled by:
  - `n.trees`: number of boosting iterations (more = risk of overfit)
  - `interaction.depth`: max tree depth (controls complexity)
  - `shrinkage`: learning rate — smaller = more conservative, needs more trees
  - `n.minobsinnode`: minimum observations per leaf — regularization
- We tune these via **5-fold cross-validation**.

```{r gbm_fit}
gbm_ctrl <- trainControl(method = "cv", number = 5)

gbm_model <- train(
  formula_stroke,
  data = train_data,
  method = "gbm",
  trControl = gbm_ctrl,
  tuneGrid = expand.grid(
    n.trees = c(2000, 2500, 3000),
    interaction.depth = c(7, 9, 11),
    shrinkage = c(0.03, 0.05, 0.07),
    n.minobsinnode = c(5, 10, 20)
  ),
  metric = "RMSE",
  verbose = FALSE
)

print(gbm_model)
plot(gbm_model, main = "GBM: CV RMSE by hyperparameters")
```
```{r gbm_importance}
gbm_imp <- summary(gbm_model$finalModel, plotit = FALSE) %>%
  as.data.frame() %>%
  rename(Feature = var, RelativeInfluence = rel.inf)

ggplot(gbm_imp, aes (x = reorder(Feature, RelativeInfluence),
                     y = RelativeInfluence)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  theme_minimal() +
  labs(title = "GBM: Relative feature influence", x = NULL, y = "Relative Influence (%)")

```

```{r gbm_predict}
gbm_preds <- predict(gbm_model, newdata = test_data)

ggplot(data.frame(obs = test_data$Stroke_m, pred = gbm_preds),
       aes(x = obs, y = pred)) +
  geom_point(alpha = 0.15, size = 0.6, color = "darkorange") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(title = "GBM: Actual vs Predicted",
       x = "Actual Stroke %", y = "Predicted Stroke %")

gbm_perf <- postResample(pred = gbm_preds, obs = test_data$Stroke_m)
gbm_perf
```

### 3. MARS — Multivariate Adaptive Regression Splines
#### Assumptions

MARS automatically searches for **breakpoints (knots)** in each predictor and fits piecewise linear functions between them. It can also detect **interaction terms** up to a specified degree. Unlike a global polynomial, each spline piece is only active in its region, making MARS more interpretable than GBM while still capturing non-linearity.

- **No linearity assumption** — MARS discovers the functional form from the data.
- **No normality or homoscedasticity assumptions** (though residual checks are still informative).
- **No feature scaling required** — knot placement is scale-invariant.
- **Independence** of observations — assumed.
- **Automatic feature selection and interaction detection** — terms that do not improve fit are pruned back.
- **Hyperparameters:**
  - `degree`: maximum interaction order. `degree = 1` = additive model (no interactions); `degree = 2` = allows pairwise interactions.
  - `nprune`: maximum number of terms retained after backward pruning.
- We tune `degree` and `nprune` via cross-validation.
```{r mars_fit}
mars_ctrl <- trainControl(method = "cv", number = 5)

mars_model <- train(
  formula_stroke,
  data = train_data,
  method = "earth",
  trControl = mars_ctrl,
  tuneGrid = expand.grid(degree = 1:2, nprune = c(5, 10, 15, 20)),
  metric = "RMSE"
)

print(mars_model)
plot(mars_model, main = "MARS: CV RMSE by degree and nprune")
```

```{r mars_summary}
summary(mars_model$finalModel)
```
```{r mars_importance}
mars_imp <- varImp(mars_model)$importance %>%
  rownames_to_column("Feature")

ggplot(mars_imp, aes(x = reorder(Feature, Overall), y = Overall)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  theme_minimal() +
  labs(title = "MARS: Variable importance", x = NULL, y = "Importance")
```

```{r mars_predict}
mars_preds <- as.numeric(predict(mars_model, newdata = test_data))

mars_plot_df <- data.frame(
  obs = test_data$Stroke_m,
  pred = as.numeric(mars_preds)
)

ggplot(mars_plot_df, aes(x = obs, y = pred)) +
  geom_point(alpha = 0.15, size = 0.6, color = "darkviolet") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(
    title = "MARS: Actual vs Predicted",
    x = "Actual Stroke %",
    y = "Predicted Stroke %"
  )
```
```{r mars_perf}
mars_perf <- postResample(pred = mars_preds, obs = test_data$Stroke_m)
mars_perf
```

```{r error_by_bin}
error_by_bin <- function(preds, obs) {
  data.frame(obs = obs, pred = preds) %>%
    mutate(
      error = pred - obs,
      abs_error = abs(error),
      bin = ntile(obs, 5)
    ) %>%
    group_by(bin) %>%
    summarise(
      n = n(),
      min_obs = min(obs),
      max_obs = max(obs),
      mean_obs = mean(obs),
      mean_pred = mean(pred),
      mean_error = mean(error),
      MAE = mean(abs_error),
      RMSE = sqrt(mean(error^2)),
      .groups = "drop"
    )
}

error_by_bin(svr_preds, test_data$Stroke_m)
```
```{r scenario2_comparison}
scenario2_results <- data.frame(
  Model = c(
    "Bayesian Linear Regression",
    "Ridge Regression",
    "Lasso Regression",
    "Support Vector Regression",
    "Gradient Boosting Regression",
    "MARS"
  ),
  Type = c(
    "Parametric",
    "Parametric",
    "Parametric",
    "Nonparametric",
    "Nonparametric",
    "Nonparametric / flexible"
  ),
  RMSE = c(
    bayes_perf["RMSE"],
    ridge_perf["RMSE"],
    lasso_perf["RMSE"],
    svr_perf["RMSE"],
    gbm_perf["RMSE"],
    mars_perf["RMSE"]
  ),
  Rsquared = c(
    bayes_perf["Rsquared"],
    ridge_perf["Rsquared"],
    lasso_perf["Rsquared"],
    svr_perf["Rsquared"],
    gbm_perf["Rsquared"],
    mars_perf["Rsquared"]
  ),
  MAE = c(
    bayes_perf["MAE"],
    ridge_perf["MAE"],
    lasso_perf["MAE"],
    svr_perf["MAE"],
    gbm_perf["MAE"],
    mars_perf["MAE"]
  )
) %>%
  mutate(
    RMSE = as.numeric(RMSE),
    Rsquared = as.numeric(Rsquared),
    MAE = as.numeric(MAE)
  ) %>%
  arrange(RMSE)

rownames(scenario2_results) <- NULL

print(scenario2_results)
```

### Scenario 2 interpretation

Scenario 2 compares parametric models with nonparametric or flexible models using the same target, predictors, preprocessing, and train/test split. If the flexible models outperform the parametric models, this supports the idea that the relationship between health predictors and stroke prevalence is not fully linear.

SVR and Gradient Boosting are expected to be the strongest predictive models. GBM is especially useful because it also provides variable importance, making it easier to interpret which predictors contribute most to the predictions.

# Overall comparison

```{r overall_comparison}
overall_results <- bind_rows(
  scenario1_results %>%
    mutate(
      Scenario = "Scenario 1",
      Type = case_when(
        Model == "Linear Regression" ~ "Parametric",
        Model == "kNN Regression" ~ "Nonparametric",
        Model == "Random Forest Regression" ~ "Nonparametric",
        TRUE ~ "Other"
      )
    ),
  scenario2_results %>%
    mutate(Scenario = "Scenario 2")
) %>%
  select(Scenario, Type, Model, RMSE, Rsquared, MAE) %>%
  arrange(RMSE)

print(overall_results)

overall_results %>%
  pivot_longer(
    cols = c(RMSE, Rsquared, MAE),
    names_to = "Metric",
    values_to = "Value"
  ) %>%
  ggplot(aes(x = reorder(Model, Value), y = Value, fill = Type)) +
  geom_col() +
  facet_wrap(~Metric, scales = "free_x") +
  coord_flip() +
  theme_minimal() +
  labs(
    title = "Overall Model Comparison",
    subtitle = "All models use the same target, selected predictors, preprocessing and train/test split",
    x = NULL,
    y = "Value",
    fill = "Model type"
  )
```

## Final discussion 

All models were trained and evaluated on the same selected feature set and the same train/test split. This makes the model comparison fair and reproducible. Hyperparameters were selected using cross-validation on the training data, while the test set was used only for final evaluation.

The results show that nonparametric and flexible models generally outperformed linear parametric models. This suggests that the relationship between stroke prevalence and the selected health predictors is not purely linear. The best-performing models were SVR and Gradient Boosting Regression.

Across several models, `Diabetes_m` and `CHD_m` were consistently the strongest predictors of stroke prevalence. This is consistent with the initial correlation analysis and with medical domain knowledge, because both variables represent strong metabolic and cardiovascular risk burden.

Linear Regression, Bayesian Linear Regression, Ridge and Lasso provided interpretable baselines. Ridge and Lasso were useful because they handle moderate multicollinearity through regularization. However, their performance was weaker than the best flexible models.

Gradient Boosting achieved strong predictive performance while also providing feature importance, making it a good balance between accuracy and interpretability. SVR achieved very strong predictive accuracy, but is less interpretable than tree-based models.

Overall, the selected feature set is justified: it excludes downstream consequences of stroke to avoid leakage and keeps upstream metabolic, cardiovascular and behavioral risk factors.

